---
title: "Conditional simulation and Rosenblatt transforms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditional simulation and Rosenblatt transforms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rvinecopulib)
set.seed(501)
```

A vine is simulated along an order. Variables to condition on must form the
tail of an admissible sampling order. rvinecopulib can enforce this during
structure selection or transiently reorient a compatible fitted model during
simulation.

## Select a conditioning-aware structure

Pass variable indices or names through `conditioning_set` when fitting a
copula. The selected structure places those variables at the end of its order.

```{r select-copula}
n <- 180
z <- rnorm(n)
x <- cbind(
  response1 = z + rnorm(n),
  response2 = -0.5 * z + rnorm(n),
  driver1 = z + rnorm(n, sd = 0.5),
  driver2 = 0.4 * z + rnorm(n)
)
u <- pseudo_obs(x)

fit <- vinecop(
  u,
  family_set = "onepar",
  conditioning_set = c("driver1", "driver2")
)
get_structure(fit)
```

Conditioning-aware selection requires a full, non-truncated vine and an MST
tree algorithm (`"mst_prim"` or `"mst_kruskal"`). Random tree algorithms do
not guarantee the required order.

## Simulate conditionally on the copula scale

`u_cond` contains one value per conditioning variable. A vector or one-row
matrix is repeated for all simulations; an `n`-row matrix specifies different
conditions for every output row.

```{r conditional-copula}
condition <- c(0.25, 0.8)
draws <- rvinecop(
  100,
  fit,
  u_cond = condition,
  conditioning_set = c("driver1", "driver2")
)
stopifnot(
  isTRUE(all.equal(draws[, "driver1"], rep(condition[1], 100))),
  isTRUE(all.equal(draws[, "driver2"], rep(condition[2], 100)))
)
head(draws)
```

When `conditioning_set` is omitted, the columns of `u_cond` refer to the last
variables in the model's current order.

```{r row-specific-conditions}
conditions <- cbind(
  driver1 = seq(0.1, 0.9, length.out = 5),
  driver2 = rep(0.5, 5)
)
row_draws <- rvinecop(
  5,
  fit,
  u_cond = conditions,
  conditioning_set = c("driver1", "driver2")
)
stopifnot(isTRUE(all.equal(row_draws[, 3:4], conditions)))
```

## Condition on the original data scale

`vine()` exposes the same structure control through `copula_controls`.
`rvine()` accepts `x_cond` on the original scale and uses the fitted margins to
compute all necessary CDF values and left limits.

```{r conditional-original-scale}
full_fit <- vine(
  as.data.frame(x),
  margins_controls = list(family_set = "kde1d"),
  copula_controls = list(
    family_set = "onepar",
    conditioning_set = c("driver1", "driver2")
  )
)

original_condition <- data.frame(driver1 = -0.5, driver2 = 1.25)
original_draws <- rvine(
  6,
  full_fit,
  x_cond = original_condition,
  conditioning_set = c("driver1", "driver2")
)
stopifnot(
  isTRUE(all.equal(original_draws[, "driver1"], rep(-0.5, 6))),
  isTRUE(all.equal(original_draws[, "driver2"], rep(1.25, 6)))
)
```

Column names are strongly recommended: they make the mapping explicit and
protect code from changes in variable order.

## Reorient an existing model transiently

Supplying an explicit `conditioning_set` to `rvinecop()` or `rvine()` asks the
backend to evaluate an equivalent orientation without modifying the model.
Only conditioning sets that can form an admissible tail of the fitted
structure are possible. If a specific set is central to the analysis, request
it during fitting; selection then guarantees a usable structure.

## Discrete conditioning variables

On the copula scale, a discrete conditioning variable needs its value `F(x)`
and left limit `F(x-)`. As with model evaluation, `u_cond` accepts:

- a compact layout with one value column per conditioning variable followed by
  one left-limit column per discrete variable; or
- an expanded layout with all value columns followed by all left-limit
  columns, where a continuous variable's two columns must be equal.

`rvine()` handles this internally from `x_cond`, including ordered and
zero-inflated margins.

## Rosenblatt transforms

The Rosenblatt transform maps observations from a fitted distribution to
independent uniforms. For continuous variables in a sampling sequence
`s1, ..., sd`, its components are
\[
Z_j = F_{s_j \mid s_1,\ldots,s_{j-1}}
\left(X_{s_j}\mid X_{s_1},\ldots,X_{s_{j-1}}\right),
\qquad j=1,\ldots,d,
\]
where the first component is unconditional. If the fitted model is correct,
the `Z_j` are independent standard uniforms. The inverse transform applies the
corresponding conditional quantiles recursively and therefore maps independent
uniforms back to the model.

```{r rosenblatt}
simulated <- rvinecop(100, fit)
independent <- rosenblatt(simulated, fit)
reconstructed <- inverse_rosenblatt(independent, fit)
max(abs(simulated - reconstructed))
```

The transform follows the model's sampling order. An explicit
`conditioning_set` requests an admissible order ending in those variables,
which is useful when conditional quantiles must follow the same orientation as
conditional simulation.

```{r conditional-rosenblatt}
transformed <- rosenblatt(
  simulated,
  fit,
  conditioning_set = c("driver1", "driver2")
)
inverse_rosenblatt(
  transformed,
  fit,
  conditioning_set = c("driver1", "driver2")
)[1:3, ]
```

For discrete variables, the transform randomizes within the jump interval by
default. See the [discrete-data guide](discrete-data.html) for the randomized
formula, deterministic upper endpoints, and required copula layouts.

## Reproducibility

Conditional simulation and randomized discrete transforms use R's
random-number state. Call `set.seed()` immediately before the operation when a
reproducible draw is required. The original model object is never mutated by a
conditional simulation or transform.

## Related documentation

- [Vine copula models and structures](vine-copula-models.html) explains the
  structure order and selection controls used here.
- [Marginal models](marginal-models.html) documents the transformation between
  original observations and the copula scale.
- The [`rvinecop()` reference](../reference/vinecop_methods.html),
  [`rvine()` reference](../reference/vine_methods.html), and [Rosenblatt
  reference](../reference/rosenblatt.html) give complete argument and return
  value details.
