---
title: "Getting started with rvinecopulib"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rvinecopulib}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rvinecopulib)
set.seed(101)
```

rvinecopulib has three main modeling interfaces. The right one depends on the
scale of the data and whether the marginal distributions are part of the model.

| Starting point | Function | Result |
|---|---|---|
| Original observations | `vine()` | Margins and a vine copula |
| Uniform pseudo-observations | `vinecop()` | A vine copula |
| Two uniform variables | `bicop()` | A bivariate copula |

The fitted classes add fit information to distribution classes. A `vine`
object is also a `vine_dist`, a `vinecop` is also a `vinecop_dist`, and a
`bicop` is also a `bicop_dist`. Simulation and evaluation therefore use the
same functions for fitted and manually constructed models.

## A complete model on the data scale

Use `vine()` when the observations are on their original scale. It estimates a
marginal distribution for every variable, transforms the data to the copula
scale, and fits the dependence model.

```{r fit-vine}
n <- 120
z <- rnorm(n)
x <- data.frame(
  amount = exp(z + rnorm(n, sd = 0.5)),
  duration = exp(0.5 * z + rnorm(n, sd = 0.8)),
  events = rpois(n, exp(0.3 + 0.25 * z))
)

fit <- vine(
  x,
  var_types = c("c", "c", "d"),
  copula_controls = list(family_set = "onepar"),
  keep_data = TRUE
)
fit
summary(fit)
```

Integer-valued discrete columns are declared with `var_types = "d"`. Ordered
factors are detected automatically. See [Marginal
models](marginal-models.html) for the default KDE margins, parametric
selection, custom margin families, zero inflation, and observation weights.

The fitted object represents the joint distribution. Its density or mass can
be evaluated on the original data scale, and new observations can be drawn
with their original column names and types.

```{r use-vine}
dvine(x[1:4, ], fit)
predict(fit, x[1:4, ], what = "pdf")
logLik(fit)
AIC(fit)

simulated <- rvine(5, fit)
simulated
```

`pvine()` and `predict(..., what = "cdf")` approximate a multivariate CDF by
Monte Carlo. Increase `n_mc` when accurate tail probabilities are important.

## A dependence model on the copula scale

Use `vinecop()` when the margins are not part of the model. Its input consists
of approximately uniform pseudo-observations. For continuous data,
`pseudo_obs()` applies scaled ranks column by column.

```{r fit-vinecop}
u <- pseudo_obs(as.matrix(USArrests))
copula_fit <- vinecop(
  u,
  family_set = c("gaussian", "clayton", "gumbel", "frank")
)
summary(copula_fit)
```

The result can be evaluated and simulated directly on the unit hypercube.

```{r use-vinecop}
new_u <- rvinecop(6, copula_fit)
dvinecop(new_u, copula_fit)
pvinecop(new_u[1:2, ], copula_fit, n_mc = 1000)
```

Do not apply ordinary ranks blindly to discrete data. Discrete copula models
need both `F(x)` and its left limit `F(x-)`; the dedicated discrete-data article
explains the accepted layouts. `vine()` constructs these quantities
automatically from its fitted margins. See [Vine copula models and
structures](vine-copula-models.html) for structure selection, truncation,
thresholding, and model inspection.

## A bivariate model

Use `bicop()` for two-dimensional copula data. It estimates each compatible
candidate and selects a family using AIC by default.

```{r fit-bicop}
u2 <- rbicop(200, "clayton", 90, 2)
bivariate_fit <- bicop(u2, family_set = "par")
bivariate_fit

dbicop(u2[1:4, ], bivariate_fit)
tail_dep(bivariate_fit)
```

The [bivariate-model guide](bivariate-copulas.html) covers rotations,
h-functions, dependence measures, family sets, vectorized parameters, and
nonparametric models.

## Fixed and fitted models

The `*_dist()` constructors are useful when the model components are known.
For example, a Gaussian copula with correlation 0.7 is a complete bivariate
distribution even though it has not been fitted:

```{r fixed-model}
fixed <- bicop_dist("gaussian", parameters = 0.7)
dbicop(c(0.25, 0.8), fixed)
rbicop(3, fixed)
```

Fitted objects additionally store a log-likelihood, observation count, and fit
controls. Set `keep_data = TRUE` only when methods such as `fitted()` need the
training data.

## Selection controls

The most frequently used controls are:

- `family_set`: candidate pair-copula families or a predefined collection;
- `selcrit`: `"loglik"`, `"aic"`, `"bic"`, `"mbic"`, or `"mbicv"` where
  applicable;
- `par_method`: maximum likelihood (`"mle"`) or Kendall's tau inversion
  (`"itau"`) for supported families;
- `allow_rotations`: whether rotated Archimedean families may be selected;
- `weights`: optional observation weights;
- `cores`: parallel workers used by supported operations.

`vinecop()` additionally controls structure selection, truncation, and
thresholding. These options are developed in the vine-model article rather
than hidden in a single large example.

## Reproducibility and persistence

Ordinary model fitting is deterministic. Simulation uses R's random-number
state, so `set.seed()` makes it reproducible. Quasi-random simulation is
available through `qrng = TRUE`.

All model objects can be stored with `saveRDS()` and restored with `readRDS()`.
Custom margin classes remain usable after restoration as long as their S3
methods are available in the R session.

```{r persistence}
path <- tempfile(fileext = ".rds")
saveRDS(copula_fit, path)
restored <- readRDS(path)
unlink(path)

set.seed(102)
draws1 <- rvinecop(4, copula_fit)
set.seed(102)
draws2 <- rvinecop(4, restored)
stopifnot(isTRUE(all.equal(draws1, draws2)))
```

## Where to go next

- [Marginal models](marginal-models.html) develops full distributions on the
  original data scale.
- [Bivariate copula models](bivariate-copulas.html) documents the pair-copula
  building blocks and family-selection controls.
- [Vine copula models and structures](vine-copula-models.html) covers
  high-dimensional dependence modeling.
- The [function reference](../reference/index.html) lists the complete API by
  modeling task.
